Le titre

### 2022-08-22 Manuscrit Thèse/chapter 3/Chapter 3.tex
taratata

\documentclass[a4paper,11pt]{report}
\usepackage[]{amsmath}
\usepackage[]{physics} % \bra, \ket etc
\usepackage{graphicx} %Pour les figures je crois
\usepackage{hyperref}
\usepackage[
    backend=biber, 
    natbib=true,
    style=numeric,
    sorting=none, %Pour faire apparaitre les refs dans l'ordre
    hyperref=true
]{biblatex} %Imports biblatex package
\addbibresource{Bib_ch3.bib} %Import the bibliography file

@@

\title{}
\begin{document}
\chapter{NV-NV cross-relaxations : the fluctuator model}
In this chapter we will look at the phenomenon of cross-relaxations between NV centers, both from an experimental and theoretical point of view. We will see how a model of fast and slow relaxing centers, the fluctuator model, can explain most of the experimental observations, and finally we will see some limitations of the model.

\section{Experimental observation of NV-NV cross-relaxation}
Before we discuss the theoretical implications of NV-NV cross-relaxations (CR), let us first show the unambiguous experimental proof of the presence of NV-NV CR in the samples with dense NV ensembles ([NV]$\gtrsim$ 1 ppm).
\subsection{NV-NV CR between nonequivalent NV centers}
\label{non_equi_valent_CR}
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{Figures/NV-NV_non_equivalent}
\caption{a) Representation of two non equivalent spins resonantly coupled. One spin is aligned with the magnetic field and therefore well polarized, while the second spin is not. b) Simulated frequencies for the $\ket{0} \to \ket{-1}$ and $\ket{0} \to \ket{+1}$ transitions when $\mathbf{B}\parallel [111]$ for the class aligned with B, and the 3 equivalent classes not aligned with B. c) Taken from \citep{armstrong2010nv}, PL change of an ensemble of NV centers as a magnetic field is scanned along the [111] axis.}
\label{non equivalent NV-NV}
\end{figure}
NV-NV CR was first observed more than thirty years ago \citep{holliday1989optical, van1989cross}. The first observations were between non-equivalent NV centers, meaning that the two NV centers involved in the dipole-dipole coupling were not polarized equally.

@@

More recently, experiments \citep{jarmola2012temperature, mrozek2015longitudinal, choi2017depolarization} on dense NV ensemble ($[\rm NV ] >1\ \rm ppm$) or at low temperature have shown that there was CR even between equivalent NV centers.

@@

Fig. \ref{equivalent NV-NV}-c) shows the transition frequencies of the $\ket{0}\to\ket{-1}$ transition for the 4 classes of NV centers on sample ADM-15-1, an HPHT micro-diamond with $[\rm NV]\sim 3\ \rm ppm$. There are three magnetic field values $B_{\rm scan} \sim$ 12, 42 and 77 G for which there is an inter-class resonance. Each of these resonances correspond to a "geometric" resonance where the $B$ field is in a symmetry plane between the two classes, meaning that the two resonant NV classes are indeed equivalent.

Finally, Fig. \ref{equivalent NV-NV}-d) shows the PL contrast as the electromagnet field is scanned. There are 3 very clear dips when $B_{\rm scan} \sim$ 12, 42 and 77 G which coincide with inter-class resonances. We will see later that this PL dip is associated with a decrease of the NV centers spin $T_1$ time from both classes.

This observation, and similar ones done by many groups, seem to indicate that there are CR between equivalent NV centers. This is however incompatible with our previous assumptions that equivalent NV centers are equally polarized and bright. To understand this phenomenon, we will need to introduce new hypotheses.

@@

\subsection{Presentation of the fluctuator model}
Choi et al. in \citep{choi2017depolarization} take this approach a step further by separating the NV centers into two groups: "normal" NV centers with a phonon-limited lifetime outside of dipole-dipole interaction, and "fluctuators" which are NV centers with an intrinsic, extremely fast relaxation mechanism. We are assuming that the fluctuator relaxation rate, $\gamma_f$ is much greater than the optical pumping rate $\Gamma_{\rm las}$. The fluctuators are therefore unpolarized and effectively act as dark spins, similarly to what we discussed in the last chapter.

@@

We will first decompose the dipole-dipole Hamiltonian in a radial and angular part:
\begin{equation}
\mathcal{H}_{\rm dd} \approx - \frac{J_0}{r^3} \left[(g+ih)(\ket{0,+1}\bra{+1,0}+\ket{0,-1}\bra{-1,0}+qS_z^1 S_z^2 \right] + h.c. ,
\end{equation}
where the expression of $J_0, g, h$ and $q$ is given in appendix [REF]. $g, h$ and $q$ are dimensionless factors that are function of the relative orientation of the two dipoles.

@@

We can note that spectral response of a single fluctuator, for which we only consider the broadening caused by the decay rate $\gamma_f$, is a Lorentzian of half width $2\gamma_f$. This is different from the usual lifetime limited optical spectra which are Lorentzian of half width $1/2T_1=\Gamma_1/2$. The difference comes here from the fact that we consider thermalization going both ways (i.e. $\ket{e} \to \ket{g}$ and $\ket{g} \to \ket{e}$ instead of only $\ket{e} \to \ket{g}$) and that each state of the fluctuator can decay into two other states.

Finally, the additional depolarization rate induced by the fluctuator on the NV center reads:
\begin{equation}
\gamma_s(\mathbf{r})=\left(\frac{J_0}{r^3}\right) \frac{\eta^2}{\gamma_f}.
\end{equation}
Compared to eq. [REF], the dependency on the inhomogeneous broadening of the spins $1/T_2^*$ is hidden in the $\eta$ factor, and we have this additional broadening $\gamma_f$ coming from the fluctuators very short lifetime.

@@

We consider each fluctuator as in independent depolarization source, therefore the depolarization on a single NV center reads: 
\begin{equation}
\gamma=\sum_i \gamma_s(\mathbf{r}_i).
\end{equation}
We then want to compute the distribution $\rho(\gamma)$ defined as:
\begin{equation}
\rho(\gamma)=\int d\{r_i\} \rho(\{r_i\})\, \delta \left( \sum_i \gamma_s(\mathbf{r}_i) - \gamma \right).
\end{equation}
To do so, we need to determine the distribution of the fluctuators positions $\{r_i\}$. Assuming that they are homogeneously distributed in the bulk of the material, we find:
\begin{equation}
\rho(\gamma)=\frac{e^{-1/(4\gamma T)}}{\sqrt{4\pi \gamma^3 T}},
\end{equation}
where the time constant $T$ was introduced and is defined as:
\begin{equation}
\frac{1}{T}=\left(\frac{4\pi n_fJ_0\bar \eta}{3}\right)^2 \frac{\pi}{\gamma_f},
\label{eq 1/T}
\end{equation}
with $n$ the fluctuator density and $\bar \eta$ the averaged value of $\abs{\eta}$:

@@

We will start by separating the factor $\eta$ given in eq. \ref{eq eta} into a a geometric and a spectral part:
\begin{align}
\eta^2&=G^2(\psi_{NV},\psi_f,\theta,\phi) R^2(\omega_f,\omega_{NV}), \\
G^2&=\abs{g}^2+\abs{h}^2,  \\ 
R^2(\omega_f,\omega_{NV}) &=\frac{1}{3} \cdot \frac{4\gamma_f^2}{(\omega_f - \omega_{NV})^2+4\gamma_f^2}.
\end{align}

@@

To compute, we need to know the distributions $\rho(\omega_{NV})$ and $\rho(\omega_{f})$. We can experimentally measure $\rho(\omega_{NV})$ though ODMR, and we can reasonably expect $\rho(\omega_{f})$ to follow the same probability.

We unfortunately cannot analytically solve $\bar R$ for standard distributions of $\rho(\omega_{f})$ and $\rho(\omega_{NV})$ (either Gaussians or Lorentzians). We can however solve $\overline{R^2}$, which is relatively close to $(\bar{R})^2$ as long as $2 \gamma_f >> 1/T_2^*$.

For a Lorentzian profile of $\rho(\omega_{f})$ and $\rho(\omega_{NV})$:
\begin{align*}
\rho(\omega_{f})&=\frac{1}{\pi \Gamma_f} \frac{1}{1+ \left(\frac{\omega_f-\omega^0_f}{\Gamma_f}\right)^2} \\
\rho(\omega_{NV})&=\frac{1}{\pi \Gamma_{NV}} \frac{1}{1+ \left(\frac{\omega_{NV}-\omega^0_{NV}}{\Gamma_{NV}}\right)^2},
\end{align*}

where $\Gamma_{f}=1/T_2^*(\rm fluctuator)$ and $\Gamma_{NV}=1/T_2^*(\rm NV)$. We then find that:

\begin{align*}
\overline{R^2}&=\frac{1}{3} \int \rho(\omega_{NV}) \rho(\omega_{f}) \frac{4\gamma_f^2}{(\omega_f - \omega_{NV})^2+4\gamma_f^2} d\omega_{NV} d\omega_{f} \\
&= \frac{1}{3}\cdot  \frac{2 \gamma_f}{2 \gamma_f + \Gamma_f + \Gamma_{NV}} \cdot \frac{1}{1+\left(\frac{\omega^0_{NV}-\omega^0_{f}}{2 \gamma_f + \Gamma_f + \Gamma_{NV}}\right)^2}.
\end{align*}

In the case where $\omega^0_{NV}=\omega^0_{f}$, meaning that either the considered NVs and fluctuators are from the same class, or from two classes perfectly resonant, by substituting $\overline{R ^2}$ for $(\bar{R})^2$ (which again is only approximately valid for $2 \gamma_f >> 1/T_2^*$) in eq. \ref{eq 1/T}, we find that:
\begin{equation}
\label{eq 1/T avec T2*}
\frac{1}{T} \propto \frac{1}{\gamma_f + 1/T_2^*},
\end{equation}
where we have assumed that $T_2^*$ had the same value for the NV centers and the fluctuators.

@@

Eq. \ref{eq 1/T avec T2*} is subject to many approximations: the inhomogeneous distribution of the NV centers is generally not Lorentzian, we don't know whether the fluctuator inhomogeneous broadening is the same than the NV one, and $\overline{R^2} \neq (\bar{R})^2$ in the general case. Nevertheless, eq. \ref{eq 1/T avec T2*} gives us a general idea of the evolution of $1/T$ with $T_2^*$.

@@

This conclusion is based on previous observations on the relaxation of nuclear spins \citep{bloembergen1949interaction, de1958relaxation, blumberg1960nuclear}, where some nuclear spins had a considerably reduced lifetime due to their proximity to paramagnetic ions, and the resonant energy transfer between molecules \citep{forster1949experimentelle, eisenthal1964influence, yokota1967effects}, where excited donor molecules are coupled through dipole-dipole interaction to non-excited acceptors, and where the stretched exponential dynamics of the donors was first observed \citep{forster1949experimentelle}.

In the case of P-doped Si, the origin of the fast relaxing centers is thought to be closely packed P impurities, so that the electronic wavefunctions of (at least) two P centers overlap. In that case two phenomena occur: the possibility of electron tunneling, and the apparition of a contact interaction term in the dipole-dipole Hamiltonian (see appendix [REF]). Both of these phenomena could lead to fast spin relaxation, the tunneling because the spin hopping is accompanied by a spin reversal \citep{sugihara1963spin}, and the modulation of the contact interaction strength $J$ by the crystal phonons \citep{honig1960electron}. Technically, the non-contact dipole-dipole interaction is also modulated by the phonons, but the effect is too small to account for the fast depolarization.

Similarly, \citep{choi2017depolarization} suggests that the origin of fluctuators in NV centers are closely packed NV centers or NV-impurity pairs which undergo rapid electron hopping and spin depolarization. To prove their point, they look at the charge dynamics in their sample and find that there is charge recombination in the dark with a corresponding tunneling rate $\sim 10\ \rm ns$. In contrast to P-doped Si, the NV centers are not the most abundant electronic defects in the crystal, this is why the NV$^--$N$^+$ pair in particular is thought to be a likely candidate for the fluctuators \citep{manson2018nv}.

@@

On the other hand, Fig. \ref{stretch_or_not_stretch}-b) shows a $T_1$ measurement coming from sample CVD-pink for B$\neq$0. This time the profile is more exponential, and the value $T_1^{\rm exp}=4.1\ \rm ms$ is coherent with a phonon-limited lifetime. Even though CVD-pink shows some NV-NV CR feature, the spin dynamics in this sample is still dominated by the phonons, which explains the exponential profile.

@@

Fig. \ref{various_fit_formulae} shows four different fitting procedure on the two same measurements. The measurements here are two $T_1$ measurements done on the same sample ADM-150-1, following the protocol described in [REF], with and without an external magnetic field ($\sim 50 G$). The magnetic field was strong enough to split the four classes and only one class was probed in the case $B\neq0$ , whereas all 4 classes were resonant in the case $B=0$, which strongly increases the NV-NV CR.

@@

We should note however that the values of $T_1^{\rm dd}$ we obtain are not absolute. Indeed, in the case of Fig. \ref{various_fit_formulae}, the value of $T_1^{\rm ph}$ can be somewhat arbitrarily fixed between 3 and 6 ms with satisfying fits, and the resulting values $T_1^{\rm dd}$ can vary by almost an order of magnitude in the case of $B\neq0$. While this technique of measuring $T_1^{\rm dd}$ is relatively sensitive and reproducible, it is not very accurate.

@@

Fig. \ref{fluct linewidth}-b) shows the dipole-dipole induced relaxation rate $\Gamma_1^{\rm dd}=T_1^{\rm dd}$ as a function of the detuning $\Delta \nu$ between the two central classes. For each magnetic field value, a $T_1$ profile was recorded and fitted following the procedure described in Fig. \ref{various_fit_formulae}-d) where we again fixed $T_1^{\rm ph}=5\ \rm ms$. We also show the ODMR spectrum of a single NV class for comparison.

We can clearly see that the $\Gamma^{\rm dd}$ profile is much broader than the ODMR line of the NV centers, whereas in a standard CR process, we expect the decay rate to be proportional to the spectral overlap of the two classes \citep{hall2016detection}. If we assume that both NV classes spectral response is a gaussian of same width $\sigma$, then the spectral overlap between the two classes is also a gaussian of width $\sqrt{2} \sigma$. Fig. \ref{fluct linewidth}-b) also shows the ODMR line scaled by a factor $\sqrt{2}$, which is still much narrower than the $1/T_1^{\rm dd}$ line.

We attribute this broadening to the fluctuators lifetime. Another potential explanation for the broadening could be the interaction strength between the NV center and the fluctuator. However, for [NV] $\sim$ 5 ppm, the average dipole-dipole coupling strength between closest NV neighbor is $\expval{\mathcal{H}_{\rm dd}} \sim 46\ \rm kHz$, which cannot explain the $\sim 4\ \rm MHz$ broadening that we observe.

@@

To do this we first assume that the dipole induced decay rate is proportional to the overlap of the spectral responses of the NV centers and fluctuators $S_{NV}$ and $S_f$. We will take $S_{NV}( \omega)$ and $S_{f}( \omega)$ as functions centered in $\omega=0$, and consider that the spectral responses are shifted by $\omega^0_f$ and $\omega^0_{NV}$ respectively, the central frequencies of each group. We then have:
\begin{align*}
\frac{1}{T_1^{\rm dd}} &\propto \int S_{NV} (\omega-\omega_{f}^0) S_f (\omega-\omega_{NV}^0) d\omega \\
&= \int S_{NV} (\omega-\Delta \omega) S_f (\omega)
 d\omega \\
 &= (S_{NV}*S_{f})(\Delta \omega),
\end{align*}

@@

We know that when $\mathbf{B}$ belongs to any of these four planes, there is a co-resonance between at least two NV classes. This co-resonance leads to an increase in NV-NV CR (or NV-fluctuator CR) and a faster depolarization of the spins involved, which ultimately leads to a drop in PL. We can indeed see in fig. \ref{Carte} that there is clear correlation between the loci of $\mathbf{B}$ and drops in PL. We can also see that the PL is at its lowest when $B\parallel [001]$, which corresponds to a four class degeneracy.

\subsection{Quantitative modification of $T_1$}
\label{sec quantitative T1}
\begin{table}[htbp]
\centering
\caption{Theoretical and experimental values of $\Gamma_1^{\rm dd}=1/T_1^{\rm dd}$ for various magnetic field configurations, expressed as a function of the value found for an isolated class. The experimental values were measured on sample ADM-15-3 using the protocol of Fig. \ref{various_fit_formulae}-d) with $T_1^{\rm ph}\equiv 5\ \rm ms$.}
 \label{T1 champ mag}
\begin{tabular}{c|cc}
\toprule
$\Gamma_1^{\rm dd} (\mathbf{B})$ &  Theory & Experimental \\
\midrule
random $\mathbf{B}$ (1 class) & $\Gamma_0^{\rm th}$ & 1.53 ms$^{-1} \equiv \Gamma_0^{\rm exp}$ \\
$\mathbf{B} \in \{110\}$ (2 classes) & 10.0 $\Gamma_0^{\rm th}$ & 5.20 $\Gamma_0^{\rm exp}$ \\
$\mathbf{B} \in \{100\}$ (2 classes) & 7.24 $\Gamma_0^{\rm th}$ & 4.22 $\Gamma_0^{\rm exp}$ \\
$\mathbf{B} \parallel \langle 111 \rangle$ (3 classes) & 28.4 $\Gamma_0^{\rm th}$ & 9.07 $\Gamma_0^{\rm exp}$ \\
$\mathbf{B} \parallel \langle 100 \rangle$ (4 classes) & 42.8 $\Gamma_0^{\rm th}$ & 11.3 $\Gamma_0^{\rm exp}$ \\
$\mathbf{B}=0$ (4 classes) & 51+($^*$) $\Gamma_0^{\rm th}$ & 15.4 $\Gamma_0^{\rm exp}$ \\
\bottomrule
\end{tabular}

>>>>

\documentclass[a4paper,11pt]{report}
\usepackage[]{amsmath}
\usepackage[]{physics} % \bra, \ket etc
\usepackage{graphicx} %Pour les figures je crois
\usepackage{hyperref}
\usepackage[
    backend=biber, 
    natbib=true,
    style=numeric-comp,
    sorting=none, %Pour faire apparaitre les refs dans l'ordre
    hyperref=true
]{biblatex} %Imports biblatex package
\addbibresource{Bib_ch3.bib} %Import the bibliography file

@@

\title{}
\begin{document}
\chapter{NV-NV cross-relaxations : the fluctuator model}
In this chapter we will look at the phenomenon of cross-relaxations between NV centers, both from an experimental and theoretical point of view.

The presence of strong NV-NV dipolar interaction in dense ensemble ([NV] $\gtrsim 1\ \rm ppm$) proved to be both an obstacle for ensemble magnetometry \citep{zhou2020quantum} and an opportunity to explore and use the many-body interaction resulting from it \citep{zhou2020quantum, choi2017observation, kucsko2018critical, zu2021emergent}.

We will see here how a model of fast and slow relaxing centers, the fluctuator model \citep{choi2017depolarization}, can explain most of the experimental observations, and we will see some limitations of the model.

\section{Experimental observation of NV-NV CR}
Before we discuss the theoretical implications of NV-NV cross-relaxations (CR), let us first show the unambiguous experimental proof of the presence of NV-NV CR in the samples with dense NV ensembles.
\subsection{NV-NV CR between nonequivalent NV centers}
\label{non_equi_valent_CR}
\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{Figures/NV-NV_non_equivalent}
\caption{a) Representation of two non equivalent spins resonantly coupled. One spin is aligned with the magnetic field and therefore well polarized, while the second spin is not. b) Simulated frequencies for the $\ket{0} \to \ket{-1}$ and $\ket{0} \to \ket{+1}$ transitions when $\mathbf{B}\parallel [111]$ for the class aligned with B, and the 3 equivalent classes not aligned with B. c) Taken from \citep{armstrong2010nv}, PL change of an ensemble of NV centers as a magnetic field is scanned along the [111] axis.}
\label{non equivalent NV-NV}
\end{figure}
NV-NV CR was first observed more than thirty years ago \citep{holliday1989optical, van1989cross}. The first observations were between non-equivalent NV centers, meaning that the two NV centers involved in the dipole-dipole coupling were not polarized equally.

@@

More recently, experiments \citep{jarmola2012temperature, mrozek2015longitudinal, jarmola2015longitudinal, choi2017depolarization} on dense NV ensemble ($[\rm NV ] >1\ \rm ppm$) or at low temperature have shown that there were cross-relaxations even between equivalent NV centers.

@@

Fig. \ref{equivalent NV-NV}-c) shows the transition frequencies of the $\ket{0}\to\ket{-1}$ transition for the 4 classes of NV centers on sample ADM-15-1, an HPHT micro-diamond with $[\rm NV]\sim 3\ \rm ppm$. There are three magnetic field values $B_{\rm scan} \approx$ 12, 42 and 77 G for which there is an inter-class resonance. Each of these resonances correspond to a ``geometric" resonance where the $B$ field is in a symmetry plane between the two classes, meaning that the two resonant NV classes are indeed equivalent.

Finally, Fig. \ref{equivalent NV-NV}-d) shows the PL contrast as the electromagnet field is scanned. There are 3 very clear dips when $B_{\rm scan} \approx$ 12, 42 and 77 G which coincide with inter-class resonances. We will see later that this PL dip is associated with a decrease of the NV centers spin $T_1$ time from both classes.

This observation, and similar ones done by many groups \citep{jarmola2012temperature, mrozek2015longitudinal, choi2017depolarization, akhmedzhanov2017microwave, giri2018coupled}, seem to indicate that there are CR between equivalent NV centers. This is however incompatible with our previous assumptions that equivalent NV centers are equally polarized and bright. To understand this phenomenon, we will need to introduce new hypotheses.

@@

\subsection{Presentation of the fluctuator model}
Choi et al. in \citep{choi2017depolarization} take this approach a step further by separating the NV centers into two groups: ``normal" NV centers with a phonon-limited lifetime outside of dipole-dipole interaction, and ``fluctuators" which are NV centers with an intrinsic, extremely fast relaxation mechanism. We are assuming that the fluctuator relaxation rate, $\gamma_f$ is much greater than the optical pumping rate $\Gamma_{\rm las}$. The fluctuators are therefore unpolarized and effectively act as dark spins, similarly to what we discussed in the last chapter.

@@

We will first decompose the dipole-dipole Hamiltonian in a radial and angular part:
\begin{equation}
\mathcal{H}_{\rm dd} \approx - \frac{J_0}{r^3} \left[(g+ih)(\ket{0,+1}\bra{+1,0}+\ket{0,-1}\bra{-1,0}+qS_z^1 S_z^2 \right] + h.c. ,
\end{equation}
where the expression of $J_0, g, h$ and $q$ is given in appendix [REF]. $g, h$ and $q$ are dimensionless factors that are function of the relative position and orientation of the two dipoles.

@@

We can note that the spectral response of a single fluctuator, for which we only consider the broadening caused by the decay rate $\gamma_f$, is a Lorentzian of half width $2\gamma_f$. This is different from the usual lifetime limited optical spectra which are Lorentzian of half width $1/2T_1=\Gamma_1/2$. The difference comes here from the fact that we consider thermalization going both ways (i.e. $\ket{e} \to \ket{g}$ and $\ket{g} \to \ket{e}$ instead of only $\ket{e} \to \ket{g}$) and that each state of the fluctuator can decay into two other states.

Finally, the additional depolarization rate induced by the fluctuator on the NV center reads:
\begin{equation}
\gamma_s(\mathbf{r})=\left(\frac{J_0}{r^3}\right) \frac{\eta^2}{\gamma_f}.
\end{equation}
Compared to eq. [REF], the dependency on the inhomogeneous broadening of the spins $1/T_2^*$ is hidden in the $\eta$ factor, and we have the additional broadening $\gamma_f$ coming from the fluctuators very short lifetime.

@@

We consider each fluctuator as in independent depolarization source, therefore the depolarization on a single NV center reads: 
\begin{equation}
\gamma=\sum_i \gamma_s(\mathbf{r}_i).
\end{equation}
We then want to compute the distribution $\rho(\gamma)$ defined as:
\begin{equation}
\rho(\gamma)=\int d\{r_i\} \rho(\{r_i\})\, \delta \left( \sum_i \gamma_s(\mathbf{r}_i) - \gamma \right).
\end{equation}
To do so, we need to determine the distribution of the fluctuators positions and orientations$\{r_i\}$. Assuming that they are homogeneously distributed in the bulk of the material, we find:
\begin{equation}
\rho(\gamma)=\frac{e^{-1/(4\gamma T)}}{\sqrt{4\pi \gamma^3 T}},
\end{equation}
where the time constant $T$ was introduced and is defined as:
\begin{equation}
\frac{1}{T}=\left(\frac{4\pi n_fJ_0\bar \eta}{3}\right)^2 \frac{\pi}{\gamma_f},
\label{eq 1/T}
\end{equation}
with $n$ the fluctuator density and $\bar \eta$ the averaged value of $\abs{\eta}$:

@@

We will start by separating the factor $\eta$ given in eq. \ref{eq eta} into a a geometric and a spectral part:
\begin{align}
\eta^2&=G^2(\psi_{NV},\psi_f,\theta,\phi) R^2(\omega_f,\omega_{NV}), \\
G^2&=\frac{1}{3}\left(\abs{g}^2+\abs{h}^2 \right),  \\ 
R^2(\omega_f,\omega_{NV}) &= \frac{4\gamma_f^2}{(\omega_f - \omega_{NV})^2+4\gamma_f^2}.
\end{align}

@@

To perform the computation, we need to know the distributions $\rho(\omega_{NV})$ and $\rho(\omega_{f})$. We can experimentally measure $\rho(\omega_{NV})$ though ODMR, and we can reasonably expect $\rho(\omega_{f})$ to follow the same probability.

We unfortunately cannot analytically solve $\bar R$ for standard distributions of $\rho(\omega_{f})$ and $\rho(\omega_{NV})$ (either Gaussians or Lorentzians). We can however solve $\overline{R^2}$, which is relatively close to $(\bar{R})^2$ as long as $2 \gamma_f \gg 1/T_2^*$.

For a Lorentzian profile of $\rho(\omega_{f})$ and $\rho(\omega_{NV})$ defined as:
\begin{align*}
\rho(\omega_{f})&=\frac{1}{\pi \Gamma_f} \frac{1}{1+ \left(\frac{\omega_f-\omega^0_f}{\Gamma_f}\right)^2} \\
\rho(\omega_{NV})&=\frac{1}{\pi \Gamma_{NV}} \frac{1}{1+ \left(\frac{\omega_{NV}-\omega^0_{NV}}{\Gamma_{NV}}\right)^2},
\end{align*}

where $\Gamma_{f}=1/T_2^*(\rm fluctuator)$ and $\Gamma_{NV}=1/T_2^*(\rm NV)$ are the inhomogeneous broadening of the NV and fluctuator ensemble, and $\omega^0_{NV}$ and $\omega^0_{f}$ are the central angular frequencies of the NV and fluctuator ensemble. We then find that:

\begin{align*}
\overline{R^2}&= \int \rho(\omega_{NV}) \rho(\omega_{f}) \frac{4\gamma_f^2}{(\omega_f - \omega_{NV})^2+4\gamma_f^2} d\omega_{NV} d\omega_{f} \\
&=\frac{2 \gamma_f}{2 \gamma_f + \Gamma_f + \Gamma_{NV}} \cdot \frac{1}{1+\left(\frac{\omega^0_{NV}-\omega^0_{f}}{2 \gamma_f + \Gamma_f + \Gamma_{NV}}\right)^2}.
\end{align*}

In the case where $\omega^0_{NV}=\omega^0_{f}$, meaning that either the considered NVs and fluctuators are from the same class, or from two classes perfectly resonant, by substituting $\overline{R ^2}$ for $(\bar{R})^2$ (which again is only approximately valid for $2 \gamma_f \gg 1/T_2^*$) in eq. \ref{eq 1/T}, we find that:
\begin{equation}
\label{eq 1/T avec T2*}
\frac{1}{T} \propto \frac{1}{2\gamma_f + \Gamma_f + \Gamma_{NV}}.
\end{equation}

@@

Eq. \ref{eq 1/T avec T2*} is subject to several approximations: the inhomogeneous distribution of the NV centers is generally not Lorentzian, and $\overline{R^2} \neq (\bar{R})^2$ in the general case. Nevertheless, it gives us a general idea of the evolution of $1/T$ with $T_2^*$.

@@

This conclusion was based on previous observations on the relaxation of nuclear spins \citep{bloembergen1949interaction, de1958relaxation, blumberg1960nuclear}, where some nuclear spins had a considerably reduced lifetime due to their proximity to paramagnetic ions, and also to the resonant energy transfer between molecules \citep{forster1949experimentelle, eisenthal1964influence, yokota1967effects}, where excited donor molecules are coupled through dipole-dipole interaction to non-excited acceptors, and where the stretched exponential dynamics of the donors was first observed \citep{forster1949experimentelle}.

In the case of P-doped Si, the origin of the fast relaxing centers is thought to be closely packed P impurities, so that the electronic wavefunctions of (at least) two P centers overlap. In that case two phenomena occur: the possibility of electron tunneling, and the apparition of a contact interaction term in the dipole-dipole Hamiltonian (see appendix [REF]). Both of these phenomena could lead to fast spin relaxation, the tunneling because the spin hopping can be accompanied by a spin reversal \citep{sugihara1963spin}, and the modulation of the contact interaction strength $J$ by the crystal phonons \citep{honig1960electron}. While the non-contact dipole-dipole interaction is also modulated by the phonons, this effect would be too small to account for the fast depolarization observed here.

Similarly, \citep{choi2017depolarization} suggests that the origin of fluctuators in NV centers are closely packed NV centers or NV-impurity pairs which undergo rapid electron hopping and spin depolarization. To prove their point, they look at the charge dynamics in their sample and find that there is charge recombination in the dark with a corresponding tunneling rate between neighboring sites of $\sim 10\ \rm ns$. In contrast to P-doped Si, the NV centers are not the most abundant electronic defects in the crystal, this is why the NV$^--$N$^+$ pair in particular is thought to be a likely candidate for the fluctuators \citep{manson2018nv}.

@@

On the other hand, Fig. \ref{stretch_or_not_stretch}-b) shows a $T_1$ measurement coming from sample CVD-pink for B$\neq$0. This time the profile is more exponential, and the value $T_1^{\rm exp}=4.1\ \rm ms$ is coherent with a phonon-limited lifetime. Even though the sample CVD-pink shows some NV-NV CR feature, the spin dynamics in this sample is still dominated by the phonons, which explains the exponential profile.

@@

Fig. \ref{various_fit_formulae} shows four different fitting procedure on the two same experimental data. The data here are two $T_1$ measurements done on the same sample ADM-150-1, following the protocol described in [REF], with and without an external magnetic field ($\sim 50 G$). The magnetic field was strong enough to split the four classes and only one class was probed in the case $B\neq0$ , whereas all 4 classes were resonant in the case $B=0$, which strongly increases the NV-NV CR.

@@

We should note however that the values of $T_1^{\rm dd}$ we obtain are not absolute. Indeed, in the case of Fig. \ref{various_fit_formulae}, the value of $T_1^{\rm ph}$ can be somewhat arbitrarily fixed between 3 and 6 ms with satisfying fits, and the resulting values $T_1^{\rm dd}$ can vary by almost an order of magnitude in the case of $B\neq0$. While this technique of measuring $T_1^{\rm dd}$ is useful to compare different cases, because it is fairly sensitive, it is not an exact measurement of the dipole-dipole induced lifetime.

@@

Fig. \ref{fluct linewidth}-b) shows the dipole-dipole induced relaxation rate $\Gamma_1^{\rm dd}=1/T_1^{\rm dd}$ as a function of the detuning $\Delta \nu$ between the two central classes. For each magnetic field value, a $T_1$ profile was recorded and fitted following the procedure described in Fig. \ref{various_fit_formulae}-d) where we again fixed $T_1^{\rm ph}=5\ \rm ms$. We also show the ODMR spectrum of a single NV class for comparison.

We can clearly see that the $\Gamma^{\rm dd}$ profile is much broader than the ODMR line of the NV centers, whereas in a standard CR process, we expect the decay rate to be proportional to the spectral overlap of the two classes \citep{hall2016detection}. If we assume that both NV classes spectral response are gaussians of same width $\sigma$, then the spectral overlap between the two classes is also a gaussian of width $\sqrt{2} \sigma$. Fig. \ref{fluct linewidth}-b) also shows the ODMR line scaled by a factor $\sqrt{2}$, which is still much narrower than the $1/T_1^{\rm dd}$ line.

We attribute this broadening to the fluctuators lifetime. Another potential explanation for the broadening could be the interaction strength between the NV center and the fluctuator, but for [NV] $\sim$ 5 ppm, the average dipole-dipole coupling strength between closest NV neighbor is $\expval{\mathcal{H}_{\rm dd}} \sim 46\ \rm kHz$, which cannot explain the $\sim 4\ \rm MHz$ broadening that we observe.

@@

To do this we first assume that the dipole induced decay rate is proportional to the overlap of the spectral responses of the NV centers and fluctuators $S_{NV}$ and $S_f$. We will take $S_{NV}( \omega)$ and $S_{f}( \omega)$ as functions centered in $\omega=0$, and consider that the spectral responses are shifted by $\omega^0_{NV}$ and $\omega^0_{f}$ respectively, the central frequencies of each group. We then have:
\begin{align*}
\frac{1}{T_1^{\rm dd}} &\propto \int S_{NV} (\omega-\omega_{NV}^0) S_f (\omega-\omega_{f}^0) d\omega \\
&= \int S_{NV} (\omega-\Delta \omega) S_f (\omega)
 d\omega \\
 &= (S_{NV}*S_{f})(\Delta \omega),
\end{align*}

@@

We know that when $\mathbf{B}$ belongs to any of these four planes, there is a co-resonance between at least two NV classes. This co-resonance leads to an increase in NV-NV CR and a faster depolarization of the spins involved, which ultimately leads to a drop in PL. We can indeed see in fig. \ref{Carte} that there is clear correlation between the loci of $\mathbf{B}$ and drops in PL. We can also see that the PL is at its lowest when $\mathbf{B}\parallel [001]$, which corresponds to a four class degeneracy.

\subsection{Quantitative modification of $T_1$}
\label{sec quantitative T1}
\begin{table}[htbp]
\centering
\caption{Theoretical and experimental values of $\Gamma_1^{\rm dd}=1/T_1^{\rm dd}$ for various magnetic field configurations, expressed as a function of the value found for an isolated class. The experimental values were measured on sample ADM-15-3 using the protocol of Fig. \ref{various_fit_formulae}-d) with $T_1^{\rm ph}= 5\ \rm ms$.}
 \label{T1 champ mag}
\begin{tabular}{c|cc}
\toprule
$\Gamma_1^{\rm dd} (\mathbf{B})$ &  Theory & Experimental \\
\midrule
random $\mathbf{B}$ (1 class) & $\Gamma_0^{\rm th}$ & 1.53 ms$^{-1} \equiv \Gamma_0^{\rm exp}$ \\
$\mathbf{B} \in \{110\}$ (2 classes) & 10.0 $\Gamma_0^{\rm th}$ & 5.20 $\Gamma_0^{\rm exp}$ \\
$\mathbf{B} \in \{100\}$ (2 classes) & 7.24 $\Gamma_0^{\rm th}$ & 4.22 $\Gamma_0^{\rm exp}$ \\
$\mathbf{B} \parallel \langle 111 \rangle$ (3 classes) & 28.4 $\Gamma_0^{\rm th}$ & 9.07 $\Gamma_0^{\rm exp}$ \\
$\mathbf{B} \parallel \langle 100 \rangle$ (4 classes) & 42.8 $\Gamma_0^{\rm th}$ & 11.3 $\Gamma_0^{\rm exp}$ \\
$\mathbf{B}=0$ (4 classes) & 51+($^*$) $\Gamma_0^{\rm th}$ & 15.4 $\Gamma_0^{\rm exp}$ \\
\bottomrule
\end{tabular}
